\bf Designing Modular Software: A Case Study in Introductory Statistics

### 2016-12-02 jcgs-paper-2016/paper.Rmd
Some more small minor edits

Modularity is a pervasive concept in computer science, extending from the design of systems
[@parnas1972criteria], to the design of software [@szyperski1996independently]. Modularity offers several advantages to both a developer and a user. In particular, functionality can be dynamically loaded and unloaded depending on the particular use case. Open source modular software opens up the possibility of extensions contributed by a wide array of programmers, which can allow the software to morph into areas that weren't anticipated early in development. In the statistics realm, R [@r-stat] is a prime example of the virtues of modular programming. As of this writing, The Comprehensive R Archive Network (CRAN) contains over 9000 source packages which can be installed and dynamically loaded in a particular session as needed.

Other statistics software also makes use of a number of these ideas. Microsoft Excel and JMP both include support for extensions, called macros and add-ins respectively, which allow programming routines to be written extending the base functionality of these programs. Compared with R, however, these programs don't maintain a large central repository of publicly available extensions on the level of CRAN. There are also software packages building upon R, and thus gaining the advantages of CRAN natively, such as R Commander [@fox2005] and Deducer [@fellows2012], which each provide a graphical front-end to many statistical functions in R. One thing these software packages all have in common is the requirement of local installation and configuration, which means certain operating systems and platforms will not support their use.

With the advantages of R clear, an approach to building statistical software and statistical learning tools would be to attempt to generate interest in programming, which could help naturally ease the transition into the use of R. Multiple software packages have recently been written in an attempt to spur this interest in R programming and statistics. DataCamp's [@datacamp] courses are user-friendly ways to learning basic R programming and data analysis techniques. Swirl [@swirl] is a similar interactive tool to make learning R more fun by learning it within R itself. Project MOSAIC [@mosaic] has created a suite of tools to simplify the teaching of statistics in the form of an R package. The primary goal of DataCamp and Swirl is to teach R programming, rather than facilitate the learning of introductory statistics.

Modern web technologies have enabled a new generation of software packages that reside solely on the web, which eliminates the issue of local installation and helps abstract away some of the more challenging programming aspects of working directly with R. Upon the release of RStudio's Shiny [@shiny] it became easier for an R-based analysis to be converted to an interactive web application. Several recent software packages have built upon Shiny to provide a web-based system based on R. One such package is iNZight Lite [@inzight] which attempts to expose students to data analysis without requiring programming knowledge. Like most web-based systems, this does not include reproducible R code, which limits its useful in a scientific or academic setting. Another package is called Radiant [@radiant], which is a web-based application with the aim of furthering business education and financial analysis. While the application is modular and extensible, it does require installation and hosting and is inundated with more features than necessary for an introductory student. An overview of the comparison between the features of these statistical software packages is presented in table \ref{tab:compare}. Partial fulfillment of requirements is noted in the table, as well as a measure of the complexity of functionality offered by default. For example, `R` does have an associated GUI, however this interface is very limited, thus only partially fulfilling the behavior of a graphical user interface.

@@

software_compare <- rbind(software_compare,
                          c("JMP", "Yes", "Yes", "Partial", "No", "Partial", "No", "No", "Full"),
                          c("R", "Partial", "Yes", "Yes", "No", "Yes", "Yes", "Yes", "Full"),
                          c("Rcmdr", "Yes", "Yes", "No", "No", "Yes", "Yes", "Yes", "Moderate"),
                          c("Deducer", "Yes", "Yes", "No", "No", "Yes", "Yes", "Yes", "Moderate"),
                          c("MOSAIC", "No", "Yes", "No", "No", "No", "Yes", "Yes", "Limited"),
                          c("iNZight Lite", "Yes", "No", "Yes", "Yes", "Partial", "No", "Yes", "Limited"),
                          c("Radiant", "Yes", "Yes", "Yes", "Yes", "Partial", "Yes", "Yes", "Moderate"))

@@

The widespread adoption of R as a tool for statistical analysis has undoubtedly been an important development for the scientific community. However, using R in most cases still requires a basic knowledge of programming concepts which may pose a steep learning curve for the introductory statistics student [@5359977]. This additional time commitment may explain why introductory courses often utilize point-and-click applications, even if the instructor himself/herself uses R in their own work. Still, some compromises must be made when using many graphical applications, including dealing with software licenses and unsupported desktop platforms. From the instructor's perspective in particular, managing a large group of software licenses for students with various computing environments and versions could wind up being extremely cumbersome.

@@

These challenges led us to create an introductory statistics application which we call `intRo`, available at http://www.intro-stats.com. `intRo` offers a number of key advantages over traditional statistics software, including ease of access and an aim to foster student interest in coding. Attempting to entirely hide the programming aspect from students, even in introductory classes, is a lost opportunity to get students interested in statistical computing. It is also a lost opportunity reaching students who learn differently or have a computational background. Another advantage is its modular structure, which allows course instructors to tailor the application towards the needs of a particular class, rather than accept a piece of software as is. Additionally, `intRo` stands apart from new tools in that it is a supplement to an existing class, fully usable by a beginning statistics student. An accompanying R package, titled `intRo` and available on GitHub, assists in the downloading, running, and deploying of `intRo` instances (See section \ref{deploying-intro-instances}).

@@

An `intRo` module is a set of self-contained executible R scripts that together produce a set of introductory statistics functionality. `intRo` modules were designed in this way to allow for simple dynamic creation of the user interface at run-time, as well as ease the process of converting existing analysis code to the `intRo` framework. A high-level diagram of this process is given in figure \ref{fig:app_creation_modules}. `intRo` modules are split up into multiple R scripts which are included either in Shiny's user interface or server definitions. At runtime, the `intRo` sources in the specified modules (contained in the modules folder) to dynamically generate the functionality available in the application. This allows for the specific functionality needed to be determined and adjusted by the individual course instructor. In this example, the instructor is electing to include a nonparametric module, which is not enabled by default, to allow the students to perform a wilcoxon rank sum test.

\begin{figure*}[ht!]
\centering
\includegraphics[width=\linewidth]{images/app_creation_modules.pdf}
\caption{This figure depicts how the Shiny \texttt{server.R} and \texttt{ui.R} files are populated using the modular structure within \texttt{intRo}. \texttt{intRo} modules are split up into multiple R scripts which are included either in Shiny's user interface or server definitions. At runtime, the \texttt{intRo} sources in the specified modules (contained in the modules folder) to dynamically generate the functionality available in the application. This allows for the specific functionality needed to be determined and adjusted by the individual course instructor. In this example, the instructor is electing to include a nonparametric module, which is not enabled by default, to allow the students to perform a wilcoxon rank sum test.}
\label{fig:app_creation_modules}
\end{figure*}

@@

The modules provided with `intRo` are contained in the modules folder. The top level directory in the modules folder defines the category of the module (currently `data`, `summaries`, or `inference`). Within each of these categories is a folder named according to the name of the module. This folder houses the previously defined scripts. As an example, we will walk through the process of creating a new module called `nonparametric`, as previously mentioned in this section, which will perform a wilcoxon rank sum test.

Since the `nonparametric` module performs a statistical test, it is part of the inference category, and hence would be placed in the `intRo` repository at **modules/inference/nonparametric**. Let's first examine *helper.R*:

@@

This script is most immediately similar to standard R code. In this case, a function `nonparametrictest` is created which, depending on the values of the parameters, ultimately returns the result of a wilcoxon rank sum test. One important difference from a typical R script is that each call in the script is wrapped in a function called `interpolate` [@interpolate]. `interpolate` both executes the given R code on the server, and also writes the code executed to the script window at the bottom of `intRo`.

Because all the code needed to implement a wilcoxon rank sum test is found in the `base` and `stats` package, the `libraries.R` file is empty for the `nonparametric` module. Additionally, there are no reactive objects defined, so `reactive.R` is also empty. `observe.R`, which defines the Shiny observers needed, is reproduced below:

@@

Reproducibility in `intRo` is accomplished with the previously mentioned `interpolate` function. `interpolate` accepts an expression and an arbitrary number of arguments as an argument, substitutes the arguments into the expression, prints the results to the console, and evaluates the parsed expression. This allows for all modules to be shoe-horned into the framework by wrapping the resulting R code in calls to `interpolate`. A possible drawback of this solution is that it requires module developers to manually wrap their functions in this call, but this could be mitigated by a package that creates modules automatically (See Section \ref{conclusions-and-future-work}).

>>>>

Modularity is a pervasive concept in computer science, extending from the design of systems
[@parnas1972criteria], to the design of software [@szyperski1996independently]. Modularity offers several advantages to both a developer and a user. In particular, functionality can be dynamically loaded and unloaded depending on the particular use case. Open source modular software precipitates the possibility of extensions contributed by a wide array of programmers, which can allow the software to morph into areas that weren't anticipated early in development. In the statistics realm, R [@r-stat] is a prime example of the virtues of modular programming. As of this writing, The Comprehensive R Archive Network (CRAN) contains over 9000 source packages which can be installed and dynamically loaded in a particular session as needed.

Other statistics software also makes use of a number of these ideas. Microsoft Excel and JMP both include support for extensions, called macros and add-ins respectively, which allow programming routines to be written extending the base functionality of these programs. Compared with R, however, these programs don't maintain a large central repository of publicly available extensions on the level of CRAN. There are also software packages building upon R, and thus gaining the advantages of CRAN natively, such as R Commander [@fox2005] and Deducer [@fellows2012], which each provide a graphical front-end to many statistical functions in R. One thing these software packages all have in common is the requirement of local installation and configuration, which means certain operating systems and platforms will not support their use. For example, Excel and JMP are not supported on Linux.

With the advantages of R clear, an approach to building statistical software and statistical learning tools would be to attempt to generate interest in programming, which could help naturally ease the transition into the use of R. Multiple software packages have recently been written in an attempt to spur this interest in R programming and statistics. DataCamp's [@datacamp] courses are a user-friendly way to learn basic R programming and data analysis techniques. Swirl [@swirl] is a similar interactive tool to make learning R more fun by learning it within R itself. Project MOSAIC [@mosaic] has created a suite of tools to simplify the teaching of statistics in the form of an R package. The primary goal of DataCamp and Swirl is to teach R programming, rather than facilitate the learning of introductory statistics.

Modern web technologies have enabled a new generation of software packages that reside solely on the web, which eliminates the issue of local installation and helps abstract away some of the more challenging programming aspects of working directly with R. Upon the release of RStudio's Shiny [@shiny] it became easier for an R-based analysis to be converted to an interactive web application. Several recent software packages have built upon Shiny to provide a web-based system based on R. One such package is iNZight Lite [@inzight] which attempts to expose students to data analysis without requiring programming knowledge. inZight Lite is currently partially but not fully reproducible. Another package is called Radiant [@radiant], which is a web-based application with the aim of furthering business education and financial analysis. While the application is modular and extensible, it does require local installation and hosting. An overview of the comparison between the features of these statistical software packages is presented in Table \ref{tab:compare}. Partial fulfillment of requirements is noted in the table, as well as a measure of the complexity of functionality offered by default. For example, `R` does have an associated Graphical User Interface (GUI), however this interface is very limited, thus only partially fulfilling the behavior of a GUI.

@@

software_compare <- rbind(software_compare,
                          c("JMP", "Yes", "Yes", "Partial", "No", "Partial", "No", "No", "Full"),
                          c("R", "Partial", "Yes", "Yes", "No", "Yes", "Yes", "Yes", "Full"),
                          c("Rcmdr", "Yes", "Yes", "No", "No", "Yes", "Yes", "Yes", "Moderate"),
                          c("Deducer", "Yes", "Yes", "No", "No", "Yes", "Yes", "Yes", "Moderate"),
                          c("MOSAIC", "No", "Yes", "No", "No", "No", "Yes", "Yes", "Limited"),
                          c("iNZight Lite", "Yes", "No", "Yes", "Yes", "Partial", "Partial", "Yes", "Limited"),
                          c("Radiant", "Yes", "Yes", "Yes", "Yes", "Partial", "Yes", "Yes", "Moderate"))

@@

The widespread adoption of R as a tool for statistical analysis has undoubtedly been an important development for the scientific community. However, using R in most cases still requires a basic knowledge of programming concepts which may pose a steep learning curve for the introductory statistics student [@5359977]. This additional time commitment may explain why introductory courses often utilize point-and-click applications, even if the instructors use R in their own work. Still, some compromises must be made when using many graphical applications, including dealing with software licenses and unsupported desktop platforms. From the instructor's perspective in particular, managing a large group of software licenses for students with various computing environments and versions could wind up being extremely cumbersome.

@@

These challenges led us to create an introductory statistics application which we call `intRo`, available at http://www.intro-stats.com. `intRo` offers a number of key advantages over traditional statistics software, including ease of access and an aim to foster student interest in coding. Attempting to entirely hide the programming aspect from students, even in introductory classes, is a lost opportunity to get students interested in statistical computing. It is also a lost opportunity reaching students who learn differently or have a computational background. Another advantage is its modular structure, which allows course instructors to tailor the application towards the needs of a particular class, rather than accept a piece of software as is. Additionally, `intRo` stands apart from new tools in that it is a supplement to an existing class, fully usable by a beginning statistics student. An accompanying R package, titled `intRo` and available on GitHub, assists in the downloading, running, and deploying of `intRo` instances (see Section \ref{deploying-intro-instances}).

@@

An `intRo` module is a set of self-contained executible R scripts that together produce a set of introductory statistics functionality. `intRo` modules were designed in this way to allow for simple dynamic creation of the user interface at run-time, as well as ease the process of converting existing analysis code to the `intRo` framework. A high-level diagram of this process is given in Figure \ref{fig:app_creation_modules}. `intRo` modules are split up into multiple R scripts which are included either in Shiny's user interface or server definitions. At runtime, the `intRo` sources in the specified modules (contained in the modules folder) to dynamically generate the functionality available in the application. This allows for the specific functionality needed to be determined and adjusted by the individual course instructor. In this example, the instructor is electing to include a nonparametric module, which is not enabled by default, to allow the students to perform a Wilcoxon rank sum test.

\begin{figure*}[ht!]
\centering
\includegraphics[width=\linewidth]{images/app_creation_modules.pdf}
\caption{This figure depicts how the Shiny \texttt{server.R} and \texttt{ui.R} files are populated using the modular structure within \texttt{intRo}. \texttt{intRo} modules are split up into multiple R scripts which are included either in Shiny's user interface (\texttt{ui.R}) or server (\texttt{helper.R}, \texttt{libraries.R}, \texttt{observe.R}, \texttt{reactive.R}, \texttt{output.R}) definitions. At runtime, \texttt{intRo} sources in the specified modules (contained in the modules folder) to dynamically generate the functionality available in the application. This allows for the specific functionality needed to be determined and adjusted by the individual course instructor. In this example, the instructor is electing to include a nonparametric module, which is not enabled by default, to allow the students to perform a Wilcoxon rank sum test.}
\label{fig:app_creation_modules}
\end{figure*}

@@

The modules provided with `intRo` are contained in the modules folder. The top level directory in the modules folder defines the category of the module (currently `data`, `summaries`, or `inference`). Within each of these categories is a folder named according to the name of the module. This folder houses the previously defined scripts. As an example, we will walk through the process of creating a new module called `nonparametric`, as previously mentioned in this section, which will perform a Wilcoxon rank sum test.

Since the `nonparametric` module performs a statistical test, it fits within the inference category, and hence should be placed in the `intRo` repository at **modules/inference/nonparametric**. Let's first create *helper.R*:

@@

This script is most immediately similar to standard R code. In this case, a function `nonparametrictest` is created which, depending on the values of the parameters, ultimately returns the result of a Wilcoxon rank sum test. One important difference from a typical R script is that each call in the script is wrapped in a function called `interpolate` [@interpolate]. `interpolate` both executes the given R code on the server, and also writes the code executed to the script window at the bottom of `intRo` (see Section \ref{designing-for-reproducibility}).

Because all the code needed to implement a Wilcoxon rank sum test is found in the `base` and `stats` package, the `libraries.R` file will be empty for the `nonparametric` module. Additionally, no reactive objects need be defined, so `reactive.R` will also be empty. `observe.R`, which defines the Shiny observers needed, can be written as follows:

@@

Reproducibility in `intRo` is accomplished with the previously mentioned `interpolate` function. `interpolate` accepts an expression and an arbitrary number of arguments as an argument, substitutes the arguments into the expression, prints the results to the console, and evaluates the parsed expression. This allows for all modules to be shoe-horned into the framework by wrapping the resulting R code in calls to `interpolate`. A possible drawback of this solution is that it requires module developers to manually wrap their functions in this call, but this could be mitigated by a package that creates modules automatically (see Section \ref{conclusions-and-future-work}).

### 2016-12-02 jcgs-paper-2016/paper.tex
Some more small minor edits

Modularity is a pervasive concept in computer science, extending from
the design of systems (Parnas 1972), to the design of software
(Szyperski 1996). Modularity offers several advantages to both a
developer and a user. In particular, functionality can be dynamically
loaded and unloaded depending on the particular use case. Open source
modular software precipitates the possibility of extensions contributed
by a wide array of programmers, which can allow the software to morph
into areas that weren't anticipated early in development. In the
statistics realm, R (R Core Team 2014) is a prime example of the virtues
of modular programming. As of this writing, The Comprehensive R Archive
Network (CRAN) contains over 9000 source packages which can be installed
and dynamically loaded in a particular session as needed.

Other statistics software also makes use of a number of these ideas.
Microsoft Excel and JMP both include support for extensions, called
macros and add-ins respectively, which allow programming routines to be
written extending the base functionality of these programs. Compared
with R, however, these programs don't maintain a large central
repository of publicly available extensions on the level of CRAN. There
are also software packages building upon R, and thus gaining the
advantages of CRAN natively, such as R Commander (Fox 2005) and Deducer
(Fellows 2012), which each provide a graphical front-end to many
statistical functions in R. One thing these software packages all have
in common is the requirement of local installation and configuration,
which means certain operating systems and platforms will not support
their use.

@@

Modern web technologies have enabled a new generation of software
packages that reside solely on the web, which eliminates the issue of
local installation and helps abstract away some of the more challenging
programming aspects of working directly with R. Upon the release of
RStudio's Shiny (RStudio and Inc. 2014) it became easier for an R-based
analysis to be converted to an interactive web application. Several
recent software packages have built upon Shiny to provide a web-based
system based on R. One such package is iNZight Lite (Wild 2015) which
attempts to expose students to data analysis without requiring
programming knowledge. inZight Lite is currently partially but not fully
reproducible. Another package is called Radiant (Nijs 2016), which is a
web-based application with the aim of furthering business education and
financial analysis. While the application is modular and extensible, it
does require installation and hosting and is inundated with more
features than necessary for an introductory student. An overview of the
comparison between the features of these statistical software packages
is presented in Table \ref{tab:compare}. Partial fulfillment of
requirements is noted in the table, as well as a measure of the
complexity of functionality offered by default. For example, \texttt{R}
does have an associated Graphical User Interface (GUI), however this
interface is very limited, thus only partially fulfilling the behavior
of a GUI.

@@

The widespread adoption of R as a tool for statistical analysis has
undoubtedly been an important development for the scientific community.
However, using R in most cases still requires a basic knowledge of
programming concepts which may pose a steep learning curve for the
introductory statistics student (Tan, Ting, and Ling 2009). This
additional time commitment may explain why introductory courses often
utilize point-and-click applications, even if the instructor
himself/herself uses R in their own work. Still, some compromises must
be made when using many graphical applications, including dealing with
software licenses and unsupported desktop platforms. From the
instructor's perspective in particular, managing a large group of
software licenses for students with various computing environments and
versions could wind up being extremely cumbersome.

@@

These challenges led us to create an introductory statistics application
which we call \texttt{intRo}, available at
\url{http://www.intro-stats.com}. \texttt{intRo} offers a number of key
advantages over traditional statistics software, including ease of
access and an aim to foster student interest in coding. Attempting to
entirely hide the programming aspect from students, even in introductory
classes, is a lost opportunity to get students interested in statistical
computing. It is also a lost opportunity reaching students who learn
differently or have a computational background. Another advantage is its
modular structure, which allows course instructors to tailor the
application towards the needs of a particular class, rather than accept
a piece of software as is. Additionally, \texttt{intRo} stands apart
from new tools in that it is a supplement to an existing class, fully
usable by a beginning statistics student. An accompanying R package,
titled \texttt{intRo} and available on GitHub, assists in the
downloading, running, and deploying of \texttt{intRo} instances (See
Section \ref{deploying-intro-instances}).

@@

An \texttt{intRo} module is a set of self-contained executible R scripts
that together produce a set of introductory statistics functionality.
\texttt{intRo} modules were designed in this way to allow for simple
dynamic creation of the user interface at run-time, as well as ease the
process of converting existing analysis code to the \texttt{intRo}
framework. A high-level diagram of this process is given in Figure
\ref{fig:app_creation_modules}. \texttt{intRo} modules are split up into
multiple R scripts which are included either in Shiny's user interface
or server definitions. At runtime, the \texttt{intRo} sources in the
specified modules (contained in the modules folder) to dynamically
generate the functionality available in the application. This allows for
the specific functionality needed to be determined and adjusted by the
individual course instructor. In this example, the instructor is
electing to include a nonparametric module, which is not enabled by
default, to allow the students to perform a wilcoxon rank sum test.

\begin{figure*}[ht!]
\centering
\includegraphics[width=\linewidth]{images/app_creation_modules.pdf}
\caption{This figure depicts how the Shiny \texttt{server.R} and \texttt{ui.R} files are populated using the modular structure within \texttt{intRo}. \texttt{intRo} modules are split up into multiple R scripts which are included either in Shiny's user interface (\texttt{ui.R}) or server (\texttt{helper.R}, \texttt{libraries.R}, \texttt{observe.R}, \texttt{reactive.R}, \texttt{output.R}) definitions. At runtime, \texttt{intRo} sources in the specified modules (contained in the modules folder) to dynamically generate the functionality available in the application. This allows for the specific functionality needed to be determined and adjusted by the individual course instructor. In this example, the instructor is electing to include a nonparametric module, which is not enabled by default, to allow the students to perform a wilcoxon rank sum test.}
\label{fig:app_creation_modules}
\end{figure*}

@@

The modules provided with \texttt{intRo} are contained in the modules
folder. The top level directory in the modules folder defines the
category of the module (currently \texttt{data}, \texttt{summaries}, or
\texttt{inference}). Within each of these categories is a folder named
according to the name of the module. This folder houses the previously
defined scripts. As an example, we will walk through the process of
creating a new module called \texttt{nonparametric}, as previously
mentioned in this section, which will perform a wilcoxon rank sum test.

@@

This script is most immediately similar to standard R code. In this
case, a function \texttt{nonparametrictest} is created which, depending
on the values of the parameters, ultimately returns the result of a
wilcoxon rank sum test. One important difference from a typical R script
is that each call in the script is wrapped in a function called
\texttt{interpolate} (Wickham 2015). \texttt{interpolate} both executes
the given R code on the server, and also writes the code executed to the
script window at the bottom of \texttt{intRo}.

Because all the code needed to implement a wilcoxon rank sum test is
found in the \texttt{base} and \texttt{stats} package, the
\texttt{libraries.R} file will be empty for the \texttt{nonparametric}
module. Additionally, no reactive objects need be defined, so
\texttt{reactive.R} will also be empty. \texttt{observe.R}, which
defines the Shiny observers needed, can be written as follows:

@@

Reproducibility in \texttt{intRo} is accomplished with the previously
mentioned \texttt{interpolate} function. \texttt{interpolate} accepts an
expression and an arbitrary number of arguments as an argument,
substitutes the arguments into the expression, prints the results to the
console, and evaluates the parsed expression. This allows for all
modules to be shoe-horned into the framework by wrapping the resulting R
code in calls to \texttt{interpolate}. A possible drawback of this
solution is that it requires module developers to manually wrap their
functions in this call, but this could be mitigated by a package that
creates modules automatically (See Section
\ref{conclusions-and-future-work}).

@@

\hypertarget{ref-r-stat}{}
R Core Team. 2014. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{http://www.R-project.org/}.

>>>>

Modularity is a pervasive concept in computer science, extending from
the design of systems (Parnas 1972), to the design of software
(Szyperski 1996). Modularity offers several advantages to both a
developer and a user. In particular, functionality can be dynamically
loaded and unloaded depending on the particular use case. Open source
modular software precipitates the possibility of extensions contributed
by a wide array of programmers, which can allow the software to morph
into areas that weren't anticipated early in development. In the
statistics realm, R (R Core Team 2016) is a prime example of the virtues
of modular programming. As of this writing, The Comprehensive R Archive
Network (CRAN) contains over 9000 source packages which can be installed
and dynamically loaded in a particular session as needed.

Other statistics software also makes use of a number of these ideas.
Microsoft Excel and JMP both include support for extensions, called
macros and add-ins respectively, which allow programming routines to be
written extending the base functionality of these programs. Compared
with R, however, these programs don't maintain a large central
repository of publicly available extensions on the level of CRAN. There
are also software packages building upon R, and thus gaining the
advantages of CRAN natively, such as R Commander (Fox 2005) and Deducer
(Fellows 2012), which each provide a graphical front-end to many
statistical functions in R. One thing these software packages all have
in common is the requirement of local installation and configuration,
which means certain operating systems and platforms will not support
their use. For example, Excel and JMP are not supported on Linux.

@@

Modern web technologies have enabled a new generation of software
packages that reside solely on the web, which eliminates the issue of
local installation and helps abstract away some of the more challenging
programming aspects of working directly with R. Upon the release of
RStudio's Shiny (RStudio and Inc. 2014) it became easier for an R-based
analysis to be converted to an interactive web application. Several
recent software packages have built upon Shiny to provide a web-based
system based on R. One such package is iNZight Lite (Wild 2015) which
attempts to expose students to data analysis without requiring
programming knowledge. inZight Lite is currently partially but not fully
reproducible. Another package is called Radiant (Nijs 2016), which is a
web-based application with the aim of furthering business education and
financial analysis. While the application is modular and extensible, it
does require local installation and hosting. An overview of the
comparison between the features of these statistical software packages
is presented in Table \ref{tab:compare}. Partial fulfillment of
requirements is noted in the table, as well as a measure of the
complexity of functionality offered by default. For example, \texttt{R}
does have an associated Graphical User Interface (GUI), however this
interface is very limited, thus only partially fulfilling the behavior
of a GUI.

@@

The widespread adoption of R as a tool for statistical analysis has
undoubtedly been an important development for the scientific community.
However, using R in most cases still requires a basic knowledge of
programming concepts which may pose a steep learning curve for the
introductory statistics student (Tan, Ting, and Ling 2009). This
additional time commitment may explain why introductory courses often
utilize point-and-click applications, even if the instructors use R in
their own work. Still, some compromises must be made when using many
graphical applications, including dealing with software licenses and
unsupported desktop platforms. From the instructor's perspective in
particular, managing a large group of software licenses for students
with various computing environments and versions could wind up being
extremely cumbersome.

@@

These challenges led us to create an introductory statistics application
which we call \texttt{intRo}, available at
\url{http://www.intro-stats.com}. \texttt{intRo} offers a number of key
advantages over traditional statistics software, including ease of
access and an aim to foster student interest in coding. Attempting to
entirely hide the programming aspect from students, even in introductory
classes, is a lost opportunity to get students interested in statistical
computing. It is also a lost opportunity reaching students who learn
differently or have a computational background. Another advantage is its
modular structure, which allows course instructors to tailor the
application towards the needs of a particular class, rather than accept
a piece of software as is. Additionally, \texttt{intRo} stands apart
from new tools in that it is a supplement to an existing class, fully
usable by a beginning statistics student. An accompanying R package,
titled \texttt{intRo} and available on GitHub, assists in the
downloading, running, and deploying of \texttt{intRo} instances (see
Section \ref{deploying-intro-instances}).

@@

An \texttt{intRo} module is a set of self-contained executible R scripts
that together produce a set of introductory statistics functionality.
\texttt{intRo} modules were designed in this way to allow for simple
dynamic creation of the user interface at run-time, as well as ease the
process of converting existing analysis code to the \texttt{intRo}
framework. A high-level diagram of this process is given in Figure
\ref{fig:app_creation_modules}. \texttt{intRo} modules are split up into
multiple R scripts which are included either in Shiny's user interface
or server definitions. At runtime, the \texttt{intRo} sources in the
specified modules (contained in the modules folder) to dynamically
generate the functionality available in the application. This allows for
the specific functionality needed to be determined and adjusted by the
individual course instructor. In this example, the instructor is
electing to include a nonparametric module, which is not enabled by
default, to allow the students to perform a Wilcoxon rank sum test.

\begin{figure*}[ht!]
\centering
\includegraphics[width=\linewidth]{images/app_creation_modules.pdf}
\caption{This figure depicts how the Shiny \texttt{server.R} and \texttt{ui.R} files are populated using the modular structure within \texttt{intRo}. \texttt{intRo} modules are split up into multiple R scripts which are included either in Shiny's user interface (\texttt{ui.R}) or server (\texttt{helper.R}, \texttt{libraries.R}, \texttt{observe.R}, \texttt{reactive.R}, \texttt{output.R}) definitions. At runtime, \texttt{intRo} sources in the specified modules (contained in the modules folder) to dynamically generate the functionality available in the application. This allows for the specific functionality needed to be determined and adjusted by the individual course instructor. In this example, the instructor is electing to include a nonparametric module, which is not enabled by default, to allow the students to perform a Wilcoxon rank sum test.}
\label{fig:app_creation_modules}
\end{figure*}

@@

The modules provided with \texttt{intRo} are contained in the modules
folder. The top level directory in the modules folder defines the
category of the module (currently \texttt{data}, \texttt{summaries}, or
\texttt{inference}). Within each of these categories is a folder named
according to the name of the module. This folder houses the previously
defined scripts. As an example, we will walk through the process of
creating a new module called \texttt{nonparametric}, as previously
mentioned in this section, which will perform a Wilcoxon rank sum test.

@@

This script is most immediately similar to standard R code. In this
case, a function \texttt{nonparametrictest} is created which, depending
on the values of the parameters, ultimately returns the result of a
Wilcoxon rank sum test. One important difference from a typical R script
is that each call in the script is wrapped in a function called
\texttt{interpolate} (Wickham 2015). \texttt{interpolate} both executes
the given R code on the server, and also writes the code executed to the
script window at the bottom of \texttt{intRo} (see Section
\ref{designing-for-reproducibility}).

Because all the code needed to implement a Wilcoxon rank sum test is
found in the \texttt{base} and \texttt{stats} package, the
\texttt{libraries.R} file will be empty for the \texttt{nonparametric}
module. Additionally, no reactive objects need be defined, so
\texttt{reactive.R} will also be empty. \texttt{observe.R}, which
defines the Shiny observers needed, can be written as follows:

@@

Reproducibility in \texttt{intRo} is accomplished with the previously
mentioned \texttt{interpolate} function. \texttt{interpolate} accepts an
expression and an arbitrary number of arguments as an argument,
substitutes the arguments into the expression, prints the results to the
console, and evaluates the parsed expression. This allows for all
modules to be shoe-horned into the framework by wrapping the resulting R
code in calls to \texttt{interpolate}. A possible drawback of this
solution is that it requires module developers to manually wrap their
functions in this call, but this could be mitigated by a package that
creates modules automatically (see Section
\ref{conclusions-and-future-work}).

@@

\hypertarget{ref-r-stat}{}
R Core Team. 2016. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{https://www.R-project.org/}.
